A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-28 latex/main.tex
Pass through dels section

% FIXME this is too detailed, and too methodsy. Abstract out and summarise
% results
Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion,
which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues
caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified
as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites''
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified
as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned

@@

\subsection{Analysis of deletion events}
\label{sec:supplement_dels}

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.
There are three contiguous deletions in NSP6
(4-base at 11288, 5-base at 11283, and 5-base at 11292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node and
the other above the BA.2 origin node in the ARG
(Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up
during deletion remapping because they had
overlapping four sites at 11288-11292
(which was mapped as the 4-base deletion event).
Furthermore, the placement of these deletion events in the ARG
reflects that of the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.

>>>>

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to disintinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?].
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, two deletions in S1-NTD [WHICH TWO???]
associated with antibody escape \cite{McCarthy2021}
See Document\ref{sec:supplement_dels} for further analysis.

caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified
as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites''
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified
as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned

@@

\subsection{Analysis of deletion events}
\label{sec:supplement_dels}
[TODO This is too terse and too vague. Be concrete and precise. WHICH 
deletions are we talking about?]

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.

[This is a separate paramgraph discussing where the model has 
broken down a bit]
There are three contiguous deletions in NSP6
(4-base at 11288, 5-base at 11283, and 5-base at 11292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node and
the other above the BA.2 origin node in the ARG
(Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up
during deletion remapping because they had
overlapping four sites at 11288-11292
(which was mapped as the 4-base deletion event).

Furthermore, the placement of these [Which events? All of them? just 
the ones we didn't properly place?] deletion events in the ARG
reflects that of the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.
